Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression

### 2020-05-19 paper/paper.tex
E -> S

  \centering
  \begin{tikzpicture}
  \newcommand\posY{0}
  \newcommand\posX{3}
  \newcommand\posE{5}
  \newcommand\posN{7}
  \newcommand\posF{10}

  \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
  \node[] (eq) at (1.5, 0){$=$};
  \node[] (times) at (4, 0){$\times$};
  \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
  \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (E) at (\posE, 0){};

  \draw[fill=white] (\posE - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posE - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
  \draw[fill=black] (\posE - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posE - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posE - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posE - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posE - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posE - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posE - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
  \draw[fill=black] (\posE - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posE - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
  \draw[fill=white] (\posE - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posE - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
  \draw[fill=black] (\posE - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posE - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);

@@

  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annE) at (\posE, -1.8){\scalebox{0.85}{$E \in \mathbb{D}^{d_x \times d_x}$}};
  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

  \node[] (labY) at (\posY, 2){Observations};
  \node[] (labX) at (\posX, 2){Factors};
  \node[] (labE) at (\posE, 2){Cause};
  \node[] (labE) at (\posE, 1.7){selection};
  \node[] (labN) at (\posN, 2){Noise};
  \node[] (labF) at (\posF, 2){Cause-effect};
  \node[] (labF) at (\posF, 1.7){mapping};

  \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
  \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
  \node[] (reg1) at (6,-3.25) {$\hat{E} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
  \end{tikzpicture}
  \caption{Back-to-back regression identifies the subset of factors $E_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
\label{fig:b2b}
\end{figure}

@@

We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{n \times
d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E
\in \mathbb{D}^{d_x \times d_x}$, we denote by $XE$ the causal factors of $Y$.
%
In summary, the problem can be formalized as:
%
\begin{equation}
    y_i = (x_i E + n_i) F
    \label{eq:model}
\end{equation}
%
where $i$ is a given sample, and $n_i$ is a sample-specific noise drawn from a
centered distribution.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of
them to have any form of covariance.
% what is a general covariance matrix?
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This problem space, along with the sizes of all variables involved, is
illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
Regression (B2B) is to estimate the matrix $E$, i.e. to identify the factors
that most reliably account for the multivariate observations.

@@

Back-to-Back Regression (B2B) consists of two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$,
and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor
of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to
$\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} =
\text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix
$E$, as detailed in the \ref{appendix:theorem_proof}.

@@

Performing two regressions over the same data sample can result in overfitting,
as spurious correlations in the data absorbed by the first regression will be
leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$
and $(X_2, Y_2)$.
%
The first regression is performed using $(X_1, Y_1)$, and the second
regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, the two
successive regressions are
repeated over many random splits, and the final estimate $\hat E$ of the causal
influence matrix is the average over the estimates associated to each split
\citep{breiman1996bagging}.
%
To accelerate this ensembling procedure, we use an efficient
leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes}
% http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf,
as follows:
%

@@

%
% After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.

@@

% JR: shall we update algorithm with LOO trick?
\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{E} \leftarrow \hat{E} + \text{diag}(\hat{H})$\;
    }
    $\hat{E} \leftarrow \hat{E} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
    \Return{$\hat{E}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centered and full
     rank noise.
     %
     Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are
     full-rank on the $Img(E)$, then, the solution of B2B, $\hat H$, will
     minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
%
\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}

Since  $E\hat H = \hat H$, we have
\begin{equation}
  \hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
\end{equation}
% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

Assuming, without loss of generality, that the active features in $E$ are the $k
\in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and
$N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:

@@

where $\Sigma_{A B}$ is the covariance of $A$ and $B$, and:
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right)
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1})
  \label{eq:diagk}
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so
$\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
Therefore, we recover $E$ from $\hat H$.

In the presence of noise, the causal factors of $E$ correspond to the positive
elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
the Appendix (\ref{recovering}).

@@

% Forward and cross-decomposition models have been used to identify the causal
% contribution of correlated factors onto multi-dimensional observations (e.g.
% \citep{naselaris2011encoding}). These approaches typically lead to multiple
% coefficients for each features (i.e. one per dimension of $Y$ or one per
% component respectively). Furthermore, these coefficients can be difficult to
% summarize into a single causal estimate. By contrast, B2B quickly (Fig.
% \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
% towards 1 and 0 for causal and non-causal features respectively.

>>>>

  \centering
  \begin{tikzpicture}
  \newcommand\posY{0}
  \newcommand\posX{3}
  \newcommand\posS{5}
  \newcommand\posN{7}
  \newcommand\posF{10}

  \node[thick, draw=black, minimum height=3cm, minimum width=2cm, fill=yellow!10] (Y) at (\posY, 0){};
  \node[] (eq) at (1.5, 0){$=$};
  \node[] (times) at (4, 0){$\times$};
  \node[thick, draw=black, minimum height=3cm, minimum width=1cm, fill=yellow!10] (X) at (\posX, 0){};
  \node[thick, draw=black, minimum height=1cm, minimum width=1cm, fill=red!10] (S) at (\posS, 0){};

  \draw[fill=white] (\posS - 0.5 + 0.0, 0.5 - 0.0) rectangle (\posS - 0.5 + 0.0 + 0.1, 0.5 - 0.0 - 0.1);
  \draw[fill=black] (\posS - 0.5 + 0.1, 0.5 - 0.1) rectangle (\posS - 0.5 + 0.1 + 0.1, 0.5 - 0.1 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.2, 0.5 - 0.2) rectangle (\posS - 0.5 + 0.2 + 0.1, 0.5 - 0.2 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.3, 0.5 - 0.3) rectangle (\posS - 0.5 + 0.3 + 0.1, 0.5 - 0.3 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.4, 0.5 - 0.4) rectangle (\posS - 0.5 + 0.4 + 0.1, 0.5 - 0.4 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.5, 0.5 - 0.5) rectangle (\posS - 0.5 + 0.5 + 0.1, 0.5 - 0.5 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.6, 0.5 - 0.6) rectangle (\posS - 0.5 + 0.6 + 0.1, 0.5 - 0.6 - 0.1);
  \draw[fill=black] (\posS - 0.5 + 0.7, 0.5 - 0.7) rectangle (\posS - 0.5 + 0.7 + 0.1, 0.5 - 0.7 - 0.1);
  \draw[fill=white] (\posS - 0.5 + 0.8, 0.5 - 0.8) rectangle (\posS - 0.5 + 0.8 + 0.1, 0.5 - 0.8 - 0.1);
  \draw[fill=black] (\posS - 0.5 + 0.9, 0.5 - 0.9) rectangle (\posS - 0.5 + 0.9 + 0.1, 0.5 - 0.9 - 0.1);

@@

  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annS) at (\posS, -1.8){\scalebox{0.85}{$S \in \mathbb{D}^{d_x \times d_x}$}};
  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

  \node[] (labY) at (\posY, 2){Observations};
  \node[] (labX) at (\posX, 2){Factors};
  \node[] (labS) at (\posS, 2){Cause};
  \node[] (labS) at (\posS, 1.7){selection};
  \node[] (labN) at (\posN, 2){Noise};
  \node[] (labF) at (\posF, 2){Cause-effect};
  \node[] (labF) at (\posF, 1.7){mapping};

  \node[] (sim1) at (0,-3.25) {\scalebox{0.85}{$X \sim P(X)$}};
  \node[] (sim2) at (0,-3.65) {\scalebox{0.85}{$N \sim P(N)$}};
  \node[] (reg1) at (6,-3.25) {$\hat{S} = \text{diag}(\underbrace{(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2\overbrace{(Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1}^{\text{1) } \hat{X} : \text{ regression from } Y \text{ to } X}}_{\text{2) regression from } X \text{ to } \hat{X}})$};
  \end{tikzpicture}
  \caption{Back-to-back regression identifies the subset of factors $S_{ii} = 1$ in $X$ that influence some observations $Y$ by 1) regressing from $Y$ to $X$ to obtain $\hat{X}$, and 2) returning the diagonal of the regression coefficients from $X$ to $\hat{X}$.}
\label{fig:b2b}
\end{figure}

@@

We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{n \times
d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $S
\in \mathbb{D}^{d_x \times d_x}$, we denote by $XS$ the causal factors of $Y$.
%
In summary, the problem can be formalized as:
%
\begin{equation}
    y_i = (x_i S + n_i) F
    \label{eq:model}
\end{equation}
%
where $i$ is a given sample, and $n_i$ is a sample-specific noise drawn from a
centered distribution.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of
them to have any form of covariance.
% what is a general covariance matrix?
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This problem space, along with the sizes of all variables involved, is
illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
Regression (B2B) is to estimate the matrix $S$, i.e. to identify the factors
that most reliably account for the multivariate observations.

@@

Back-to-Back Regression (B2B) consists of two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$,
and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor
of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to
$\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{S} =
\text{diag}(\hat{H})$, is the desired estimate of the causal influence matrix
$S$, as detailed in the \ref{appendix:theorem_proof}.

@@

Performing two regressions over the same data sample can result in overfitting,
as spurious correlations in the data absorbed by the first regression will be
leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ into two splits $(X_1, Y_1)$
and $(X_2, Y_2)$.
%
The first regression is performed using $(X_1, Y_1)$, and the second
regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, the two
successive regressions are
repeated over many random splits, and the final estimate $\hat S$ of the causal
influence matrix is the average over the estimates associated to each split
\citep{breiman1996bagging}.
%
To accelerate this ensembling procedure, we use an efficient
leave-one-out cross-validation scheme as detailed in \citep{rifkin2007notes}
% http://cbcl.mit.edu/publications/ps/MIT-CSAIL-TR-2007-025.pdf,
as follows:
%

@@

%
% After obtaining $\hat{S}$, we can fit a final regression from $X \hat{S}$ to $Y$.

@@

% JR: shall we update algorithm with LOO trick?
\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{S} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{S} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{S} \leftarrow \hat{S} + \text{diag}(\hat{H})$\;
    }
    $\hat{S} \leftarrow \hat{S} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{S}, Y)$\;
    \Return{$\hat{S}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

     Consider the B2B model from Equation $Y = (XS + N)F$, $N$ centered and full
     rank noise.
     %
     Let $Img(M)$ refers to the image of the matrix $M$. If $F$ and $X$ are
     full-rank on the $Img(S)$, then, the solution of B2B, $\hat H$, will
     minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $S\hat H = \hat H$
\end{theorem}
%
\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}

Since  $S\hat H = \hat H$, we have
\begin{equation}
  \hat H = \arg \min_H  \left \| X - XSH\right\| ^2  + \left \| NSH\right \| ^2 = (S X^\top XS +SN^\top NS) ^\dagger SXX^\top.
\end{equation}
% So, $\hat H = \arg \min_H  \left \| X - XSH\right\| ^2  + \left \| NSH\right \| ^2 = (S X^\top XS +SN^\top NS) ^\dagger SXX^\top$.

Assuming, without loss of generality, that the active features in $S$ are the $k
\in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and
$N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:

@@

where $\Sigma_{A B}$ is the covariance of $A$ and $B$, and:
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right)
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1})
  \label{eq:diagk}
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so
$\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(S)$$
Therefore, we recover $S$ from $\hat H$.

In the presence of noise, the causal factors of $S$ correspond to the positive
elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
the Appendix (\ref{recovering}).

@@

% Forward and cross-decomposition models have been used to identify the causal
% contribution of correlated factors onto multi-dimensional observations (e.g.
% \citep{naselaris2011encoding}). These approaches typically lead to multiple
% coefficients for each features (i.e. one per dimension of $Y$ or one per
% component respectively). Furthermore, these coefficients can be difficult to
% summarize into a single causal estimate. By contrast, B2B quickly (Fig.
% \ref{fig:duration}) leads to a single unbiased scalar values $\hat S$ tending
% towards 1 and 0 for causal and non-causal features respectively.

### 2020-05-19 paper/synthetic.tex
E -> S

The data generating process for each experiment constructs $n=1000$ training examples
according to the model $Y = (\text{h} XE + N)F$, where $\text{h}$ is a
scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{n \times d_x}$
%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $E \in
%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
%entries drawn from $\mathcal{N}(0, \sigma^2)$, \item $\Sigma_N = BB^\top$,
%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
%$\mathcal{N}(0, \sigma^2)$, \end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
$X \in \mathbb{R}^{n \times d_x}$ contains rows
drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $E \in \mathbb{R}^{d_x
\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
\times d_x}$ contains entries drawn from $\mathcal{N}(0, \sigma^2)$, and the
factor $\text{h} \in \mathcal{R}_+$.

@@

%
B2B leads to \emph{scalar} coefficients for non-causal
features. The diagonal of this matrix, $\hat E$, can thus be directly used
as a causal contribution estimate. Note that this estimate is unbiased (i.e.
zeros-centered) only if the second regression $H$ is not regularized.

In contrast, the Forward, Backward, CCA and PLS models lead to a
loading \emph{vector} for each feature: each vector indicates how to a feature
relates to each dimension of the observations (for Forward or Backward
models) or each canonical projection of the observations.
To estimate causal contribution, we must thus transform such vector into a
scalar, by e.g. taking the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}
Note that in such B2B cases, the estimates are  thus necessarily positive and thus biased.

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup, but not in the neuroimaging experiments below.

@@

In most real-world cases, $E$ is not known. Consequently, the above AUC evaluation can not be estimated.
To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
each feature (for the B2B and Backward models) for each dimension of $Y$ (for
the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.

>>>>

The data generating process for each experiment constructs $n=1000$ training examples
according to the model $Y = (\text{h} XS + N)F$, where $\text{h}$ is a
scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{n \times d_x}$
%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $S \in
%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
%entries drawn from $\mathcal{N}(0, \sigma^2)$, \item $\Sigma_N = BB^\top$,
%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
%$\mathcal{N}(0, \sigma^2)$, \end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
$X \in \mathbb{R}^{n \times d_x}$ contains rows
drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $S \in \mathbb{R}^{d_x
\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
\times d_x}$ contains entries drawn from $\mathcal{N}(0, \sigma^2)$, and the
factor $\text{h} \in \mathcal{R}_+$.

@@

%
B2B leads to \emph{scalar} coefficients for non-causal
features. The diagonal of this matrix, $\hat S$, can thus be directly used
as a causal contribution estimate. Note that this estimate is unbiased (i.e.
zeros-centered) only if the second regression $H$ is not regularized.

In contrast, the Forward, Backward, CCA and PLS models lead to a
loading \emph{vector} for each feature: each vector indicates how to a feature
relates to each dimension of the observations (for Forward or Backward
models) or each canonical projection of the observations.
To estimate causal contribution, we must thus transform such vector into a
scalar, by e.g. taking the sum of square coefficients:
% \begin{equation}
  $\hat S_i = \sum_j {H^j_i}^2 $
% \end{equation}
Note that in such B2B cases, the estimates are  thus necessarily positive and thus biased.

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(S, \hat S)$.
%\begin{equation} AUC = 1 - \sum_1^n (S_k - S_{k-1}) ( \hat{S}_k +
%\hat{S}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup, but not in the neuroimaging experiments below.

@@

In most real-world cases, $S$ is not known. Consequently, the above AUC evaluation can not be estimated.
To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-one feature $X_{-i}$ (i.e. 'knock-out X'). This
procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$ for
each feature (for the B2B and Backward models) for each dimension of $Y$ (for
the Forward model) or each canonical dimension of $Y$ (for CCA and PLS). The
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each
$X_i$ improves the prediction of (a pre-defined dimension or projection of) $Y$.
